Type Evolution in Programming Systems

### 2024-05-28 paper.tex

\frameworkbox{
\begin{wrapfigure}{r}{9em}
\includegraphics[scale=0.2]{figures/diag-variants.png}
\end{wrapfigure}
\frameworkboxtitle{Primitive Type Transformations}
In this example, \emph{type} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@

\newpage
~
\newpage
~
\newpage

@@

Acme Corp has a problem, for spreadsheets don't naturally support relationships between entities.
End-user databases like Airtable~\cite{airtable} and Notion~\cite{notion} support \textit{links}
between tables but do not provide operations that can do an \textit{Extract Entity}. Neither do
the established database products (see Related Work). Acme will need to write code or do a big manual conversion.

TODO: Maybe insert something about methodology in general?
(ref the table challenges paper? although this paper is no longer so much about challenges)

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases, which provide surprisingly little help.
SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is still mostly done by writing custom SQL code to alter the schema and migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.

>>>>

\begin{figure}[H]
\begin{tikzcd}[sep = 6em]
  \textsf{type}_1
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_1"{name=C, description}]
  &
  \textsf{data}_1
  \arrow[d, "\textsf{migrate}_1"{name=R, description}]
  \\
  \textsf{type}_2
  \arrow[r, "\textsf{constrains}"{above}]
  \arrow[d, red, "\textsf{change}_2"{name=C2, description}]
  &
  \textsf{data}_2
  \arrow[Rightarrow, from=C, to=R, "\textsf{propagate}"{above}]
  \arrow[d, "\textsf{migrate}_2"{name=R2, description}]
  \\
  \textsf{type}_3
  \arrow[r, "\textsf{constrains}"{above}]
  &
  \textsf{data}_3
  \arrow[Rightarrow, from=C2, to=R2, "\textsf{propagate}"{above}]
\end{tikzcd}
\caption{Multiversion type evolution.}
\label{fig:multiversion}
\end{figure}

\frameworkbox{
% wrapfigure doesn't seem to work with tikz
% \begin{wrapfigure}{r}{9em}
% \includegraphics[scale=0.2]{figures/diag-variants.png}
% \end{wrapfigure}

\frameworkboxtitle{Primitive Type Transformations}
In this example, \textsf{type} in Figure~\ref{fig:multiversion} refers to data tables, but the same scenario applies to
objects or JSON values in other programming systems. Our example focuses on permanent \emph{data}
that need to evolve alongside with the type and keep synchronized even in the non-local scenario.
The evolution can be done through a small set of primitive type transformations (illustrated on the
right) with corresponding data transformation for each. The system may either expose the
primitive transformations through a user interface, or infer them from textual edits made by the
programmer, e.g. to a database table schema definition.
}

@@

\newpage
~
\newpage

@@

\textcolor{red}{Maybe insert something about methodology in general?
(ref the table challenges paper? although this paper is no longer so much about challenges)}

There has been a lot of research on evolving interdependent layers of software, much of it guided by schema and types. But the vast majority of these techniques involve writing custom code in some specialized language, rather than through immediate action in an interactive programming system. Nevertheless it is important to understand these batch techniques to envision more interactive ones.

\paragraph{Schema evolution.}
Schema evolution has long been a major problem for SQL databases because
the SQL Data Definition Language (DDL) has limited capabilties to redefine existing data. It can rename tables and columns. SQLite \cite{sqliteDatatypes} uses dynamically typed values allowing them to be implicitly converted if the datatype of the column changes.
MYSql \cite{mysqlAlterTable} can reorder columns without destroying their data. But apart from such special cases, SQL databases have no general purpose support for data migration. As a result in practice schema evolution is mostly done by writing custom Data Manipulation Language (DML) code to migrate the data. Such custom code is greatly complicated if it needs to be done without taking the database offline or must be coordinated across multiple shards.
In \emph{Refactoring Databases} \citet{ambler06} offer a comprehensive taxonomy of schema evolution patterns, including typical strategies for online migration and sample SQL implementations.
